---
title: "Advanced Density Visualization with ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Density Visualization with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.align = "center"
)
```

While the `sshist` package provides native `plot()` S3 methods for rapid exploratory data analysis, modern analytical workflows frequently rely on `ggplot2` for creating publication-ready visualizations.

This vignette demonstrates how to extract the structural outputs from all six primary functions of the package and transform them into tidy data frames, and construct advanced plots using `ggplot2` and `patchwork`

```{r setup, message=FALSE, warning=FALSE}
library(sshist)
options(sshist.ncores = 2)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) {
  library(ggplot2)
  library(patchwork)
  theme_set(theme_minimal(base_size = 12))
}
```

---

## 1. Univariate (1D) Visualizations

We use the classic `faithful` dataset to explore univariate distributions, focusing on the `waiting` time variable.

### 1.1 Optimal 1D Histogram

The `sshist()` function returns an optimal set of breakpoints (`edges`). Instead of manually constructing rectangles, the most robust way to replicate the optimal histogram in `ggplot2` is to pass these edges directly to the `breaks` argument of `geom_histogram()`. 

*Note: By explicitly providing `breaks`, we bypass ggplot2's default alignment and `pretty()` binning logic, ensuring the histogram is mathematically exact and perfectly anchored to the data's true domain.*

```{r ggplot-sshist, fig.height = 4, eval = has_ggplot2}
data(faithful)

# 1. Calculate optimal binning parameters
res_hist <- sshist(faithful$waiting)

# 2. Plot using raw data but mapping to the optimized breaks
ggplot(faithful, aes(x = waiting)) +
  geom_histogram(
    breaks = res_hist$edges,
    fill = "steelblue",
    color = "white",
    alpha = 0.85,
    aes(y = after_stat(density))
  ) +
  geom_rug(alpha = 0.5, color = "slategrey") +
  labs(
    title = "Optimal 1D Histogram (Shimazaki-Shinomoto)",
    subtitle = sprintf("Optimized bins: %d | Bin width: %.2f", res_hist$opt_n, res_hist$opt_d),
    x = "Waiting time (minutes)",
    y = "Density"
  )
```

### 1.2 Fixed vs. Adaptive 1D Kernel Density (`sskernel` & `ssvkernel`)

For kernel estimations, the functions return evaluation coordinates `x` and density values `y`. We can combine a fixed global bandwidth estimator and a locally adaptive estimator to compare how they handle data.

A "heavy" visualization pattern is to show the density estimate on top.

```{r ggplot-kde-1d, fig.height = 4, eval = has_ggplot2}
# Define a shared grid for perfect alignment
shared_grid <- seq(min(faithful$waiting), max(faithful$waiting), length.out = 500)

# Run fixed and adaptive estimators
res_fixed <- sskernel(faithful$waiting, tin = shared_grid)
res_adaptive <- ssvkernel(faithful$waiting, tin = shared_grid)

# Create a tidy data frame for density lines
df_density <- data.frame(
  time      = c(res_fixed$x, res_adaptive$x),
  density   = c(res_fixed$y, res_adaptive$y),
  Estimator = rep(c("Fixed Global", "Locally Adaptive"), each = length(shared_grid))
)

# Visualize the comparison
ggplot(df_density, aes(x = time, y = density, color = Estimator)) +
  geom_line(linewidth = 1) +
  geom_area(aes(fill = Estimator), alpha = 0.1, position = "identity") +
  scale_color_manual(values = c("Fixed Global" = "#4575b4", "Locally Adaptive" = "#d73027")) +
  scale_fill_manual(values = c("Fixed Global" = "#4575b4", "Locally Adaptive" = "#d73027")) +
  labs(
    title = "1D KDE Comparison: Fixed vs. Adaptive",
    x = "Waiting time (minutes)", 
    y = "Density"
  ) +
  theme(legend.position = "top")
```

### 1.3 Comparing Window Functions (`WinFunc`) in Adaptive KDE

The `ssvkernel` function uses Nadaraya-Watson kernel regression to smooth the locally adaptive bandwidths over time. The `WinFunc` parameter determines the shape of the weighting function used in this secondary smoothing step. 

We can visually compare how different analytical window functions ("Boxcar", "Gauss", "Laplace", "Cauchy") influence the final adaptive density estimate.

```{r ggplot-winfunc-comparison, fig.height = 4, eval = has_ggplot2}
# Run adaptive estimators with all available window functions
res_boxcar  <- ssvkernel(faithful$waiting, tin = shared_grid, WinFunc = "Boxcar")
res_gauss   <- ssvkernel(faithful$waiting, tin = shared_grid, WinFunc = "Gauss")
res_laplace <- ssvkernel(faithful$waiting, tin = shared_grid, WinFunc = "Laplace")
res_cauchy  <- ssvkernel(faithful$waiting, tin = shared_grid, WinFunc = "Cauchy")

# Create a tidy data frame for plotting
df_winfunc <- data.frame(
  time    = rep(shared_grid, 4),
  density = c(res_boxcar$y, res_gauss$y, res_laplace$y, res_cauchy$y),
  Window  = factor(
    rep(c("Boxcar", "Gauss", "Laplace", "Cauchy"), each = length(shared_grid)),
    levels = c("Boxcar", "Gauss", "Laplace", "Cauchy")
  )
)

# Visualize the effect of different weight functions
ggplot(df_winfunc, aes(x = time, y = density, color = Window)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "Effect of Window Functions on Adaptive KDE",
    subtitle = "Comparing Boxcar, Gauss, Laplace, and Cauchy weight distributions",
    x = "Waiting time (minutes)",
    y = "Density"
  ) +
  theme(legend.position = "right")
```

---

## 2. Bivariate (2D) Visualizations

For 2D space, we analyze `eruptions` duration and `waiting` time simultaneously.

### 2.1 Optimal 2D Histogram (`sshist_2d`)

`sshist_2d()` calculates the mathematically ideal grid partitions for X and Y independently. In `ggplot2`, we can use `stat_bin2d` and supply the optimal number of bins or bin widths calculated by the algorithm to ensure object-driven plotting.

```{r ggplot-sshist2d, fig.width = 6, fig.height = 6, eval = has_ggplot2}
res_hist2d <- sshist_2d(faithful$eruptions, faithful$waiting)

# Extract optimal bin counts or widths if explicit
# Since ggplot2 handles bins natively, we supply the calculated breaks or counts
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  stat_bin2d(
    breaks = list(
      x = seq(min(faithful$eruptions), max(faithful$eruptions), length.out = res_hist2d$opt_nx + 1L),
      y = seq(min(faithful$waiting), max(faithful$waiting), length.out = res_hist2d$opt_ny + 1L)
    ),
    color = "black",
    linewidth = 0.2
  ) +
  scale_fill_distiller(palette = "YlGnBu", direction = -1) +
  labs(
    title = "Optimal 2D Histogram (Old Faithful)",
    subtitle = sprintf("Grid: %d \u00d7 %d bins", res_hist2d$opt_nx, res_hist2d$opt_ny),
    x = "Eruption duration (min)",
    y = "Waiting time (min)",
    fill = "Count"
  )
```

### 2.2 Fixed vs. Adaptive 2D Kernel Density (`sskernel2d` & `ssvkernel2d`)

Our 2D KDE functions return optimized matrices (`z`). We can unravel these matrices into a tidy `x, y, z` format using `expand.grid()` to leverage `geom_raster()`.

Let's run both `sskernel2d` (Fixed) and `ssvkernel2d` (Adaptive) and display them side-by-side using `patchwork` to visualize the exact differences in smoothing patterns.

```{r ggplot-kde2d, fig.width = 11, fig.height = 5.5, eval = has_ggplot2}
# Compute densities on a 150x150 evaluation grid
grid_res <- 150
res_2d_fixed <- sskernel2d(faithful$waiting, faithful$eruptions, n_grid = grid_res)
res_2d_adap  <- ssvkernel2d(faithful$waiting, faithful$eruptions, n_grid = grid_res)

# Helper function to wrangle the list output into a tidy data frame
wrangle_2d <- function(obj) {
  df <- expand.grid(x = obj$x_grid, y = obj$y_grid)
  df$density <- as.vector(obj$z) 
  return(df)
}

df_fixed <- wrangle_2d(res_2d_fixed)
df_adap  <- wrangle_2d(res_2d_adap)

# Plot 1: Fixed Isotropic KDE
p1 <- ggplot(df_fixed, aes(x = x, y = y, fill = density)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(option = "mako") +
  labs(title = "Fixed Global Bandwidth (sskernel2d)", x = "Waiting time (min)", y = "Eruption duration (min)") +
  theme(legend.position = "none")

# Plot 2: Locally Adaptive KDE
p2 <- ggplot(df_adap, aes(x = x, y = y, fill = density)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(option = "mako") +
  labs(title = "Locally Adaptive Bandwidth (ssvkernel2d)", x = "Waiting time (min)", y = "Eruption duration (min)", fill = "Density")

# Combine side by side with a shared legend
p1 + p2 + plot_layout(guides = "collect")
```

The locally adaptive estimator (`ssvkernel2d`) demonstrably tightens the density around the concentrated clusters and smooths the vacant regions, avoiding the "halo" effect typical of fixed bandwidth constraints.

```{r fallback, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("**Note:** The figures in this vignette require the **ggplot2** and **patchwork** packages, which are not available in this environment. Install them with `install.packages(c(\"ggplot2\", \"patchwork\"))` and rebuild the vignette.")
```
